\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2015-09-15 nnt_manu.Rnw
more about package structure, need to trim the fat

% knitr options
<<setup,include = F,cache = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, digits = 1)
@
% get online bib file
<<echo = FALSE, cache = FALSE>>=
refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
refs <- rawToChar(refs$content)
writeLines(refs, con = file('refs.bib'))
@

@@

The following is a brief description of the basic structure and methods for creating a network to provide sufficient context for the \pkg{NeuralNetTools} package.  A detailed discussion of the \ac{mlp} neural network is beyond the scope of this manuscript and interested readers should consult \citet{Rumelhart86,Ripley96} and references therein for a more thorough review.  An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.  Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.  The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables. `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  `Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

@@

Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}. An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a relatively new package that represents less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models should be comparable.  Differences between the packages primarily include default values of the arguments and the ability to handle multiple response variables or multiple hidden layers in the same model.  The \code{nnet} function can take separate (or combined) input or response inputs as each as a separate \code{data.frame} or as a \code{formula}, the \code{neuralnet} function can only use a \code{formula} as input, and the \code{mlp} function (\pkg{RSNNS}) can only take a \code{data.frame} as combined or separate variables as input.  The \code{neuralnet} function is not capable of modelling multiple response variables, unless the response is a categorical variable where one node for each outcome is used in the response.  Additionally, the default output for the \code{neuralnet} function is linear, whereas the opposite is true for the other two functions.  Other differences that influence how functions in \pkg{NeuralNetTools} handle object classes from the different packages are described below.

@@

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks.  Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).

This function plots a neural network as a neural interpretation diagram as in \citet{Ozesmi99}. Options to plot without color-coding or shading of weights are also provided. The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to relative magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Neural networks created using \code{mlp} do not show bias layers.

>>>>

% knitr options
<<setup,include = F,cache = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, digits = 1)
@
% get online bib file
<<echo = FALSE, cache = FALSE>>=
refs <- httr::GET('https://raw.githubusercontent.com/fawda123/refs/master/refs.bib')
refs <- rawToChar(refs$content)
writeLines(refs, con = file('refs.bib'))
@

@@

An intriguing characteristic of the neural network approach to predictive modelling is the absence of assumptions regarding the distributional characteristics of the response variables.  Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.  The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables. `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  `Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

@@

Packages available in \proglang{R} to create \ac{mlp} neural networks include \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}. The \pkg{train} can be used to create a neural network model by repeated training with the \code{nnet}method available from \pkg{nnet}.  An additional package, \pkg{FCNN4R}, provides an interface between \pkg{R} and the \pkg{FCNN} \proglang{C++} library.  At the time of writing, \pkg{FCNN4R} is a new package with less than 1\% of the total downloads of all neural network packages on CRAN.  As such, \pkg{NeuralNetTools} does not include methods for the \pkg{FCNN4R} package, although future versions may do so depending on the popularity of \pkg{FCNN4R}.  In general, the remaining packages use similar methods to create \ac{mlp} networks such that the resulting models are comparable.  Differences between the packages primarily include default values of the arguments and the ability to include multiple response variables or multiple hidden layers.  The primary functions in each package also vary in the type of input used to specify the model structure.  Differences that influence how the \pkg{NeuralNetTools} functions handle object classes from the different packages are described below.

@@

The stable release of \pkg{NeurelNetTools} can be installed from CRAN and the development version can be installed from GitHub:

<<eval = F>>=
# install from CRAN
install.packages('NeuralNetTools')
library(NeuralNetTools)

# install from GitHub
install.package('devtools')
library(devtools)
install_github('fawda123/NeuralNetTools')
library(NeuralNetTools)
@

\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.  These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.  A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.  The function is used internally within the main package functions but may be useful for comparing networks of different object classes.

A sample dataset is also included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  The \code{neuraldat} package provides data following this structure.  Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.  Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.

Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics. A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.  The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to relative magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.

<<echo = F>>=
library(NeuralNetTools)
@
<<>>=
# create a model using the nnet package
# neuraldat is included with NeuralDat
@

### 2015-09-15 nnt_manu.tex
more about package structure, need to trim the fat

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks.  Although useful for viewing the basic structure, the output is visually minimal and does not include options for customization (verify).

This function plots a neural network as a neural interpretation diagram as in \citet{Ozesmi99}. Options to plot without color-coding or shading of weights are also provided. The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to relative magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Neural networks created using \code{mlp} do not show bias layers.

>>>>

The stable release of \pkg{NeurelNetTools} can be installed from CRAN and the development version can be installed from GitHub:

\begin{kframe}
\begin{alltt}
\hlcom{# install from CRAN}
\hlkwd{install.packages}\hlstd{(}\hlstr{'NeuralNetTools'}\hlstd{)}
\hlkwd{library}\hlstd{(NeuralNetTools)}

\hlcom{# install from GitHub}
\hlkwd{install.package}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/NeuralNetTools'}\hlstd{)}
\hlkwd{library}\hlstd{(NeuralNetTools)}
\end{alltt}
\end{kframe}

\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.  These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.  A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.  The function is used internally within the main package functions but may be useful for comparing networks of different object classes.

A sample dataset is also included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 300 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}.  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  A common approach for data preprocessing prior to creating a nueral network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  The \code{neuraldat} package provides data following this structure.  Novel datasets can be preprocessed in the common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.  Although useful for viewing the basic structure, the output is visually minimal and does not include extensive options for customization.

Tbe \code{plotnet} function plots a neural network as a neural interpretation diagram \citep{Ozesmi99} and includes many options to customize plot aesthetics. A neural interpration diagram is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to change the thickness and color of the weight connections based on mangnitude and sign, respecitively.  The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to relative magnitude of each weight. The first layer includes only input variables with nodes labelled arbitrarily as I1 through In for n input variables. One through many hidden layers are plotted with each node in each layer labelled as H1 through Hn. The output layer is plotted last with nodes labeled as O1 through On. Bias nodes connected to the hidden and output layers are also shown. Differences in the behavior of \code{plotnet} for different object classes are minimal, with the exception of \code{mlp} class networks that do not include bias layers.

\begin{kframe}
\begin{alltt}
\hlcom{# create a model using the nnet package}
\hlcom{# neuraldat is included with NeuralDat}
\end{alltt}
\end{kframe}
